---
title: "Dethlefsen and Relman Microbiome Dataset"
author: "Daniel J. Park"
date: "10/25/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(dplyr)
```

Dethlefsen and Relman: 
- three individuals (D, E, F)
- 52 to 56 stool samples per subject
- over 10 months
- two 5-day courses (external perturbations) of the antibiotic "ciprofloxacin" (Cp) separated by 6 months
- samples were collected daily for two 19-day periods surrounding each Cp course and weekly or monthly at other times

Taxa are rows; Samples are columns
```{r}
dethlefsen.relman.df <- read.delim(file = "/Users/danieljbpark/Downloads/dethlefsen_relman_ds_01.txt")
```

We don't have metadata in a file format!!! 

```{r}
dim(dethlefsen.relman.df)

dr.D.df <- dethlefsen.relman.df %>%
  select(matches("D\\d", ignore.case = FALSE))

dr.E.df <- dethlefsen.relman.df %>%
  select(matches("E\\d", ignore.case = FALSE))

dr.F.df <- dethlefsen.relman.df %>%
  select(matches("F\\d", ignore.case = FALSE))

dim(dr.D.df); dim(dr.E.df); dim(dr.F.df)


dethlefsen.relman.df %>%
  select(D1, Name, Acc_No, Full_Name, Strain, Domain, Phylum, Taxon_3, Taxon_8) %>%
  head(10)
```