---
title: "pldist halfvarson analysis"
author: "Anna Plantinga"
date: "4/15/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
library(ggbeeswarm)
library(lme4)
library(lmerTest)
library(pldist)
library(MiRKAT)
library(vegan)
library(tidyverse)
library(phyloseq)
library(microbiome)
library(gridExtra)
source("./mp_analysis_functions.R")
```


Read in data: 

```{r} 
metadata <- read.delim("Halfvarson 2017 IBS data/Halfvarson_2017_Metadata.txt")
otus.orig <- read.delim("Halfvarson 2017 IBS data/Halfvarson_2017_OtuTable.txt")
load("Halfvarson 2017 IBS data/Halfvarson_2017_Tree.Rda")
```

Prepare OTU table: 

  - Filter to samples with at least 10,000 reads 
  - Then filter to OTUs that appear in more than 2 samples 

```{r} 
## Total read counts of at least 10000 
otus.rc <- otus.orig[, which(apply(otus.orig, 2, sum) >= 10000)]

## 9178 OTUs => 6588 OTUs 
num.nz <- apply(otus.rc, 1, FUN = function(x) sum(x!=0))
otus.nz <- otus.rc[which(num.nz > 2), ] 

## Keep only these sample IDs in metadata 
metadata2 <- metadata %>% 
  filter(sample_name %in% colnames(otus.nz))
```

Restrict to at least 4 time points: 

  - 516 samples belonging to 84 patients 
  - new.metadata: 504 samples belonging to 83 patients
    - 8 HC, 75 IBD
AND no missing BMI or f-cal
  - 462 samples belonging to 77 patients

```{r} 
# when subjects have more than two samples at a time point, choose one arbitrarily 
tab.tp.id <- table(metadata2$patientnumber, metadata2$timepoint) 
problem.IDS <- names(which(apply(tab.tp.id, 1, 
      FUN = function(x) sum(x > 1)) > 0))
which.dup <- apply(
  with(subset(metadata2, patientnumber %in% problem.IDS), 
     table(patientnumber, timepoint)), 
  1, FUN = function(x) which(x > 1))
excl.dups <- c() 
for (i in 1:length(which.dup)) {
  this.patient <- names(which.dup)[i] 
  this.time <- which.dup[i]
  this.samps <- which(metadata2$patientnumber == as.numeric(this.patient) & 
          metadata2$timepoint == this.time)
  excl.dups <- c(excl.dups, this.samps[2:length(this.samps)])
}

metadata2 <- metadata2[-excl.dups, ]

metadata3 <- metadata2 %>%
  mutate(bmi = as.numeric(bmi), 
         calprotectin = as.numeric(calprotectin)) %>%
  filter(!is.na(bmi) & !is.na(calprotectin))

# choose subjects with 4 or more time points with BMI and f-cal
subjID <- metadata3 %>%
  group_by(patientnumber) %>%
  summarise(n = n()) %>%
  filter(n >= 4) %>%
  pull(patientnumber)
subjID2 <- metadata3 %>%
  filter(ibd_subtype != "HC") %>%
  group_by(patientnumber) %>%
  summarise(n = n()) %>%
  filter(n >= 4) %>%
  pull(patientnumber)

n.totalsubj <- length(subjID2)

# the new subjects' metadata
new.metadata <- metadata3 %>%
  filter(patientnumber %in% subjID) 

new.metadata2 <- metadata3 %>%
  filter(patientnumber %in% subjID2) 

new.metadata2 %>% 
  group_by(ibd_subtype) %>% 
  distinct(patientnumber) %>% 
  summarise(n = n())

# the new subjects' samples
new.otutab <- otus.nz[ , new.metadata$sample_name]
rarefied.otutab <- t(rrarefy(t(new.otutab), min(colSums(new.otutab))))

new.otutab2 <- otus.nz[ , new.metadata2$sample_name]
```

# PLDIST

```{r}
otus.pl <- t(new.otutab2)
meta.pl <- data.frame(subjID = new.metadata2$patientnumber, 
                      sampID = new.metadata2$sample_name, 
                      time = as.numeric(new.metadata2$timepoint))
plD <- pldist_all(otus = otus.pl, 
           metadata = meta.pl, 
           paired = FALSE, 
           clr = FALSE, 
           method = c("b", "u"), 
           tree = hvs.tree)

Ks <- lapply(plD, FUN = function(d) D2K(d))

plD.patientnumber <- data.frame(patientnumber = as.integer(colnames(plD$D_braycurtis_qual)))

outcome.df <- new.metadata2 %>%
  group_by(patientnumber) %>%
  slice(1) %>%
  select(patientnumber, ibd_subtype)

outcome.df <- left_join(plD.patientnumber, outcome.df, by = "patientnumber")
# outcome.df <- outcome.df %>%
#   mutate(y = if_else(ibd_subtype == "HC", 1, 0))

X = new.metadata2 %>% 
  group_by(patientnumber) %>%
  summarise(bmi = mean(bmi),
            calprotectin = mean(calprotectin), 
            sex = first(sex)) %>%
  mutate(bmi = scale(bmi),
         calprotectin = scale(calprotectin))

# sink("halfvarson-results/halfvarson-pldist3.txt")
# print(MiRKAT(y = outcome.df$y, X = X, Ks = Ks))
# print(names(Ks))
# sink()
```

```{r}
library(ape)

pcoa_brayB <- pcoa(plD$D_braycurtis_qual)

brayB.df <- as.data.frame(pcoa_brayB$vectors) %>%
  mutate(y = outcome.df$y)

pcoa_brayplot <- ggplot(brayB.df, aes(x = Axis.1, y = Axis.2)) +
  geom_point(aes(color = as.factor(y))) +
  labs(title = "PCoA of pldist-transformation distances", 
       subtitle = "Binary Bray-Curtis dissimilarity",
       x = "Axis 1", y = "Axis 2", color = "IBD Status") +
  scale_color_manual(labels = c("ICD subgroups", "UC subgroups"),
                     values = c("red", "turquoise")) +
  theme(legend.position = "bottom")

pcoa_unifracU <- pcoa(plD$D_UW)

uniU.df <- as.data.frame(pcoa_unifracU$vectors) %>%
  mutate(y = outcome.df$y)

pcoa_uniplot <- ggplot(uniU.df, aes(x = Axis.1, y = Axis.2)) +
  geom_point(aes(color = as.factor(y))) +
  labs(title = "PCoA of pldist-transformation distances", 
       subtitle = "Unweighted Unifrac Distance",
       x = "Axis 1", y = "Axis 2", color = "IBD Status") +
  scale_color_manual(labels = c("ICD subgroups", "UC subgroups"),
                     values = c("red", "turquoise")) +
  theme(legend.position = "none")

g_legend <- function(a.gplot) {
  tmp <- ggplot_gtable(ggplot_build(a.gplot)) 
  leg <- which(sapply(tmp$grobs, function(x) x$name) == "guide-box")
  legend <- tmp$grobs[[leg]]
  return(legend)
}

shared.legend <- g_legend(pcoa_brayplot)
pcoa_brayplot <- pcoa_brayplot + theme(legend.position = "none")

par(mar = c(5.1, 4.1, 4.1, 2.1))
grid.arrange(arrangeGrob(pcoa_brayplot, pcoa_uniplot, ncol=2, nrow = 1), 
             arrangeGrob(shared.legend, ncol=1, nrow=1), 
             nrow = 2, 
             heights = c(4, 1))

```

# LMVoltest

1. read count btwn groups
2. OTU distribution btwn groups

```{r}
# QUAL LMVoltest
rel.otutab <- transform(otu_table(new.otutab2, taxa_are_rows = TRUE), transform = "compositional")
rel.otutab.rfd <- transform(otu_table(rarefied.otutab, taxa_are_rows = TRUE), transform = "compositional")

otu.avg.abnd <- apply(rel.otutab, 1, mean)       # global OTU average relative abundances
otu.quintiles <- data.frame(otu.avg.abnd) %>%     # abundance-based OTU quintiles
  rownames_to_column(var = "otu.id") %>%
  mutate(quintile = ntile(otu.avg.abnd, n = 5)) 

OTUDist.df <- data.frame(PropUnderThresh = numeric(n.totalsubj),
                         PatientType = character(n.totalsubj))
thresh <- 0.0001

input_1 <- list() # group 1: otutab and subject ID
input_2 <- list() # group 2
for (i in 1:n.totalsubj) {
  subj.id <- subjID2[i]
  subj.metadata <- new.metadata2 %>%
    filter(patientnumber == subj.id) %>%
    mutate(date = as.Date(collection_timestamp, format = "%m/%d/%Y")) %>%
    arrange(date)
    
  subj.otutab <- as.data.frame(rel.otutab.rfd[ , subj.metadata$sample_name])
  subj.avg.abnd <- apply(subj.otutab, 1, mean)
  subj.p <- mean(subj.avg.abnd < thresh)
  if("UC" %in% subj.metadata$ibd_subtype) {
    input_1[[length(input_1)+1]] <- list(subj.otutab, subj.id)
    OTUDist.df[i ,] <- c(subj.p, "HC")
  }
  else {
    input_2[[length(input_2)+1]] <- list(subj.otutab, subj.id)
    OTUDist.df[i ,] <- c(subj.p, "IBD")
  } 
}

big.statechange.tab_1 <- statechange_prop_table(input_1) %>%
  mutate(outcome = 0) # UC patients
big.statechange.tab_1 <- left_join(big.statechange.tab_1, X %>% rename(subj.id = patientnumber), 
                                   by = c("subj.id"))

big.statechange.tab_2 <- statechange_prop_table(input_2) %>%
  mutate(outcome = 1) # CCD and ICD patients
big.statechange.tab_2 <- left_join(big.statechange.tab_2, X %>% rename(subj.id = patientnumber), 
                                   by = c("subj.id"))

big.statechange.tab <- rbind(big.statechange.tab_1, big.statechange.tab_2)
big.statechange.tab <- left_join(x = big.statechange.tab, y = otu.quintiles, by = "otu.id")
big.statechange.tab <- big.statechange.tab %>%
  mutate(bmi = scale(bmi),
         calprotectin = scale(calprotectin),
         quintile = as.factor(quintile))

glmm <- glmer(prop.statechange ~ 1 + outcome + bmi + calprotectin + sex + quintile + (1|subj.id) + (1|otu.id),
              data = big.statechange.tab, 
              family = binomial, 
              weights = timepoints,
              verbose = 1,
              control = glmerControl(optCtrl = list(maxfun = 1000)))

glmm3 <- glmer(prop.statechange ~ 1 + outcome + bmi + calprotectin + sex + (1|subj.id) + (1|otu.id),
               data = big.statechange.tab,
               family = binomial,
               weights = timepoints,
               verbose = 1,
               control = glmerControl(optCtrl = list(maxfun = 1000)))

ss <- getME(glmm, c("theta", "fixef"))
glmm2 <- update(glmm, start = ss, control = glmerControl(optCtrl = list(maxfun = 2e4)))

summary(glmm2)
fe <- coef(glmm2)
re <- ranef(glmm2)

sink("halfvarson-results/lmvoltest-qual2.txt")
print(summary(glmm2))
sink()
```

```{r}
group2coef <- coef(summary(glmm2))["outcome", 1]

subjRE <- re$subj.id %>%
  mutate(patientnumber = as.integer(rownames(re$subj.id)))

RE.df <- left_join(outcome.df, subjRE, by = "patientnumber")
RE.df <- RE.df %>% 
  mutate(intercept = if_else(ibd_subtype == "HC", `(Intercept)`, `(Intercept)` + group2coef))

lmvoltest.qual.plot <- ggplot(RE.df, aes(x = as.factor(y), y = intercept)) +
  geom_beeswarm(aes(color = as.factor(y)), cex = 4) +
  labs(title = "LMVoltest Subject-Specific \n Random Intercepts", 
       subtitle = "Adjusted by Group Fixed Effect Coefficient",
       x = "", color = "IBD status") +
  scale_color_manual(labels = c("IBD", "Healthy Control"),
                     values = c("#0072B2", "#D55E00")) +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        legend.position = "bottom")

OTUDist.df$PropUnderThresh <- as.numeric(OTUDist.df$PropUnderThresh)
ggplot(OTUDist.df, aes(x = PatientType, y = PropUnderThresh)) +
  geom_boxplot()

x <- OTUDist.df %>% filter(PatientType == "IBD") %>% select(PropUnderThresh)
y <- OTUDist.df %>% filter(PatientType == "HC") %>% select(PropUnderThresh)
t.test(x, y)

readcounts <- colSums(new.otutab)
RC.metadata <- new.metadata %>%
  select(sample_name, ibd_subtype)
RC.metadata$ReadCounts <- readcounts[RC.metadata$sample_name]
RC.metadata$PatientType <- if_else(RC.metadata$ibd_subtype == "HC", "HC", "IBD")
ggplot(RC.metadata, aes(x = PatientType, y = log(ReadCounts))) +
  geom_boxplot()

wilcox.test(x = RC.metadata %>% filter(PatientType == "IBD") %>% pull(ReadCounts),
            y = RC.metadata %>% filter(PatientType == "HC") %>% pull(ReadCounts))
```


```{r}
# QUANT LMVoltest
big.logFC.tab <- data.frame()

for (i in 1:n.totalsubj) {
  subj.id <- subjID[i]
  subj.metadata <- new.metadata %>%
    filter(patientnumber == subj.id) %>%
    arrange(timepoint)
  subj.otutab <- as.data.frame(new.otutab[ , subj.metadata$sample_name])
  sim.logFC.tab <- fold_difference_table(subj.otutab, logRatio = TRUE)
  # within-sample squared L2 norm of log fold-changes excluding dis/reappearances and continued absences
  logFC_i <- apply(sim.logFC.tab, 2, FUN = function(x) mean((x[is.finite(x)])**2, na.rm = TRUE))
  
  if("HC" %in% subj.metadata$ibd_subtype) {
    outcome = rep(0, length(logFC_i))
  } else {
    outcome = rep(1, length(logFC_i))
  }
  
  subj.df <- data.frame(logFC = logFC_i,
                        outcome = outcome,
                        bmi = subj.metadata$bmi[2:(length(logFC_i)+1)],
                        calprotectin = subj.metadata$calprotectin[2:(length(logFC_i)+1)],
                        sex = rep(subj.metadata$sex[1], length(logFC_i)),
                        subj.id = rep(subj.id, length(logFC_i)))
  big.logFC.tab <- rbind(big.logFC.tab, subj.df)
}

big.logFC.tab <- big.logFC.tab %>%
  mutate(bmi = scale(bmi),
         calprotectin = scale(calprotectin))

lmm <- lmer(logFC ~ 1 + outcome + bmi + calprotectin + sex + (1|subj.id), 
            data = big.logFC.tab)

re <- ranef(lmm)

sink("halfvarson-results/lmvoltest-quant2.txt")
print(summary(lmm))
sink()
```

```{r}
group2coef <- coef(summary(lmm))["outcome", 1]

subjRE <- re$subj.id %>%
  mutate(patientnumber = as.integer(rownames(re$subj.id)))

RE.df <- left_join(outcome.df, subjRE, by = "patientnumber")
RE.df <- RE.df %>% 
  mutate(outcome = if_else(ibd_subtype == "HC", "HC", "IBD"),
         intercept = if_else(ibd_subtype == "HC", `(Intercept)`, `(Intercept)` + group2coef))

lmvoltest.quant.plot <- ggplot(RE.df, aes(x = as.factor(y), y = intercept)) +
  geom_beeswarm(aes(color = as.factor(y)), cex = 4) +
  labs(title = "LMVoltest Subject-Specific \n Random Intercepts", 
       subtitle = "Adjusted by Group Fixed Effect Coefficient",
       x = "", color = "IBD status") +
  scale_color_manual(values = c("#0072B2", "#D55E00")) +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        legend.position = "none")

g_legend <- function(a.gplot) {
  tmp <- ggplot_gtable(ggplot_build(a.gplot)) 
  leg <- which(sapply(tmp$grobs, function(x) x$name) == "guide-box")
  legend <- tmp$grobs[[leg]]
  return(legend)
}

shared.legend <- g_legend(lmvoltest.qual.plot)
lmvoltest.qual.plot <- lmvoltest.qual.plot + theme(legend.position = "none")

par(mar = c(5.1, 4.1, 4.1, 2.1))
grid.arrange(arrangeGrob(lmvoltest.qual.plot, lmvoltest.quant.plot, ncol=2, nrow = 1), 
             arrangeGrob(shared.legend, ncol=1, nrow=1), 
             nrow = 2, 
             heights = c(4, 1))
```

# ADHOC

```{r}
## create data dataset with: distance, subj.id, group2
## binary (qualitative) and quantitative distances
big.dist.tab.qual <- data.frame()
big.dist.tab.quant <- data.frame()
for (i in 1:n.totalsubj) {
  subj.id <- subjID[i]
  subj.metadata <- new.metadata %>%
    filter(patientnumber == subj.id) %>%
    arrange(timepoint)
  subj.otutab <- as.data.frame(new.otutab[ , subj.metadata$sample_name])
  
  n.pairs <- ncol(subj.otutab) - 1 # number of consecutive pairs of samples
  
  if ("HC" %in% subj.metadata$ibd_subtype) {
    group2 <- rep(0, n.pairs) # not in group2 if subject is healthy
  } else {
    group2 <- rep(1, n.pairs) # in group2 if subject has IBD
  }
  
  dist.mat.qual <- as.matrix(vegdist(x = t(subj.otutab), method = "bray", binary = TRUE)) # sample-sample distance matrix
  dist.tab.qual_i <- data.frame(distance = diag(dist.mat.qual[2:nrow(dist.mat.qual), 1:(nrow(dist.mat.qual)-1)]),
                                group2 = group2,
                                bmi = subj.metadata$bmi[2:(n.pairs+1)],
                                calprotectin = subj.metadata$calprotectin[2:(n.pairs+1)],
                                sex = rep(subj.metadata$sex[1], length(n.pairs)),
                                subj.id = rep(subj.id, n.pairs))
  dist.mat.quant <- as.matrix(vegdist(x = t(subj.otutab), method = "bray", binary = FALSE)) # sample-sample distance matrix
  dist.tab.quant_i <- data.frame(distance = diag(dist.mat.quant[2:nrow(dist.mat.quant), 1:(nrow(dist.mat.quant)-1)]),
                                 group2 = group2,
                                bmi = subj.metadata$bmi[2:(n.pairs+1)],
                                calprotectin = subj.metadata$calprotectin[2:(n.pairs+1)],
                                sex = rep(subj.metadata$sex[1], length(n.pairs)),
                                 subj.id = rep(subj.id, n.pairs))
  ## rbind with data dataset
  big.dist.tab.qual <- rbind(big.dist.tab.qual, dist.tab.qual_i)
  big.dist.tab.quant <- rbind(big.dist.tab.quant, dist.tab.quant_i)
}

big.dist.tab.qual <- big.dist.tab.qual %>%
  mutate(bmi = scale(bmi),
         calprotectin = scale(calprotectin))
big.dist.tab.quant <- big.dist.tab.quant %>%
  mutate(bmi = scale(bmi),
         calprotectin = scale(calprotectin))

library(glmmTMB)
# LMM: distance ~ 1 + group2 + (1|subj.id)
lmm.qual <- lmer(distance ~ 1 + group2 + bmi + calprotectin + sex + (1|subj.id), 
                 data = big.dist.tab.qual)
# lmm.qual <- glmmTMB(distance ~ 1 + group2 + bmi + calprotectin + sex + (1|subj.id), 
#                     data = big.dist.tab.qual, 
#                     family = beta_family)
sink("halfvarson-results/bray-qual3.txt")
print(summary(lmm.qual))
sink()

lmm.quant <- lmer(distance ~ 1 + group2 + bmi + calprotectin + sex + (1|subj.id), 
                  data = big.dist.tab.quant)
# lmm.quant <- glmmTMB(distance ~ 1 + group2 + (1 | subj.id), 
#                     data = big.dist.tab.quant, 
#                     family = beta_family)
sink("halfvarson-results/bray-quant3.txt")
print(summary(lmm.quant))
sink()
```

```{r}
brayplot_qual <- ggplot(big.dist.tab.qual, aes(x = as.factor(group2), y = distance)) +
  geom_boxplot(aes(fill = as.factor(group2))) +
  labs(title = "Intra-subject Pairwise Dissimilarities", x = "", y = "Bray-Curtis dissimilarity (binary)",
       fill = "IBD subgroup") +
  scale_fill_manual(labels = c("ICD", "UC"),
                     values = c("red", "turquoise")) +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        legend.position = "bottom")

brayplot_quant <- ggplot(big.dist.tab.quant, aes(x = as.factor(group2), y = distance)) +
  geom_boxplot(aes(fill = as.factor(group2))) +
  labs(title = "Intra-subject Pairwise Dissimilarities", x = "", y = "Bray-Curtis dissimilarity (quantitative)",
       fill = "IBD subgroup") +
  scale_fill_manual(labels = c("ICD", "UC"),
                     values = c("red", "turquoise")) +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        legend.position = "none")

g_legend <- function(a.gplot) {
  tmp <- ggplot_gtable(ggplot_build(a.gplot)) 
  leg <- which(sapply(tmp$grobs, function(x) x$name) == "guide-box")
  legend <- tmp$grobs[[leg]]
  return(legend)
}

shared.legend <- g_legend(brayplot_qual)
brayplot_qual <- brayplot_qual + theme(legend.position = "none")

par(mar = c(5.1, 4.1, 4.1, 2.1))
grid.arrange(arrangeGrob(brayplot_qual, brayplot_quant, ncol=2, nrow = 1), 
             arrangeGrob(shared.legend, ncol=1, nrow=1), 
             nrow = 2, 
             heights = c(4, 1))
```
