---
title: "Moving Pictures Magnitude Reappearance"
author: "Daniel J. Park"
date: "7/16/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(ggplot2)
library(dplyr)
library(zoib)
require(foreign)
require(Hmisc)
source("mp_analysis_functions.R")
```

```{r}
# f4.metadata : metadata sorted by time
# otu.counts.f4 : OTU counts
# otu.relabs.f4 : OTU relative abundances
# otu.clrabs.f4 : OTU clr-transformed abundances
# median.otu.relabs.f4 : OTU's median relative abundance across all samples
# median.otu.clrabs.f4 : OTU's median clr-transformed abundance across all samples
# coverage.f4 : the total number of reads by sample
load("~/Downloads/mp_F4_data.Rdata")
load("~/Downloads/mp_M3_data.Rdata")

```

# Magnitude of reappearance

Set up a regression problem for predicting the magnitude of a reappearing OTU count (the positive count for OTU $k$ at time $t$ given a zero count for OTU $k$ at time $t-1$).

```{r}
mag.reapp.f4 <- format_reappearance_data(otutab = otu.counts.f4, 
                                         t.sample = f4.metadata$days_since_experiment_start, 
                                         otu.median.abs = median.otu.clrabs.f4,
                                         read.depths = coverage.f4,
                                         otu_id = names(median.otu.clrabs.f4))

head(mag.reapp.f4); tail(mag.reapp.f4); summary(mag.reapp.f4); dim(mag.reapp.f4)

mag.reapp.f4
```


For a pretty rare OTU, is there a difference in the distributions of sample depth (total read count) for reappearance and non-reappearance samples?

```{r}
rare.otus <- c("174950", "183207", "495451")
median.otu.relabs.f4[which(names(median.otu.relabs.f4) == "174950" )]
median.otu.relabs.f4[which(names(median.otu.relabs.f4) == "183207" )]
median.otu.relabs.f4[which(names(median.otu.relabs.f4) == "495451" )]

rare.reapps <- subset(mag.reapp.f4, otu_ids %in% rare.otus)
dim(rare.reapps)

rare.reapps <- rare.reapps %>% 
  mutate(isReappearance = case_when(y == 0 ~ "non-reappearance", y > 0 ~ "reappearance"))

rare.reapps

ggplot(rare.reapps) +
  geom_boxplot(aes(x = isReappearance, y = sample.read.depth, fill = as.factor(otu_ids)))
 
```


The frequency table looks like the distribution of reappearance counts is highly zero-inflated.

```{r}
ggplot(mag.reapp.f4, aes(y)) +
  geom_histogram(binwidth = 1)

ggplot(mag.reapp.f4, aes(x=median.abundance, y)) +
  geom_point()

ggplot(mag.reapp.f4, aes(x=sample.read.depth, y)) +
  geom_point()

ggplot(mag.reapp.f4, aes(x=time.diff, y)) +
  geom_point()

as.data.frame(table(mag.reapp.f4$y))
```


0) (Zero-inflated?) Negative Binomial Regression

```{r}
require(pscl)
require(MASS)
require(boot)
```

A first pass zero-inflated negative binomial model predicting the reappearance count.

```{r}
m1 <- zeroinfl(y ~ median.abundance | median.abundance,
               data = mag.reapp.f4, dist = "negbin")

summary(m1)
m1$coefficients
m1$theta
```

Compared to a model with just an intercept, m1 is statistically significant.
```{r}
m0 <- update(m1, . ~ 1)

pchisq(2 * (logLik(m1) - logLik(m0)), df = 2, lower.tail=FALSE)
```

Model 2 is not zero-inflated.

```{r}
m2 <- glm.nb(y ~ median.abundance, data = mag.reapp.f4)

summary(m2)
m2$coefficients
m2$theta
```


Compare the zero-inflated vs. not-inflated negative binomial regression models. Is the Vuong test an appropriate method?
```{r}
vuong(m1, m2)
```

Add time between consecutive samples into the mix, but only an intercept for zero-inflation binomial with logit link.
```{r}
m3 <- zeroinfl(y ~ median.abundance + time.diff | 1,
               data = mag.reapp.f4, dist = "negbin")

summary(m3)
```

The negative binomial part depends on abundance ranking and log sample depth.

```{r}
m4 <- zeroinfl(y ~ median.abundance + log(sample.read.depth) | 1,
               data = mag.reapp.f4, dist = "negbin")

summary(m4)
```


No zero-inflation, but include all the covariates in the negative binomial regression.

```{r}
m5 <- glm.nb(y ~ median.abundance + log(sample.read.depth) + time.diff, data = mag.reapp.f4)

summary(m5)
```

Compare predictions from zero-inflated vs. not-inflated negative binomial regression models.

```{r}
new.data <- mag.reapp.f4 %>% dplyr::select( median.abundance, sample.read.depth, time.diff)

new.data$pred.m1 <- predict_reappearance(X_count = list(new.data$median.abundance), 
                                         X_zero = list(new.data$median.abundance), 
                                         negbin_model = m1, zero_inflated = TRUE, seed=0)

new.data$pred.m2 <- predict_reappearance(X_count = list(new.data$median.abundance), 
                                         negbin_model = m2, zero_inflated = FALSE, seed=0)

new.data$pred.m3 <- predict_reappearance(X_count = list(new.data$median.abundance, new.data$time.diff), 
                                         negbin_model = m3, zero_inflated = TRUE, seed=0)

new.data$pred.m4 <- predict_reappearance(X_count = list(new.data$median.abundance, log(new.data$sample.read.depth)),
                                         negbin_model = m4, zero_inflated = TRUE, seed=0)

new.data$pred.m5 <- predict_reappearance(X_count = list(new.data$median.abundance, log(new.data$sample.read.depth), new.data$time.diff),
                                         negbin_model = m5, zero_inflated = FALSE, seed=0)
```

```{r}
ggplot(mag.reapp.f4)+
  geom_point(aes(x = median.abundance, y = y), color="purple") +
  labs(x = "OTU median clr-transformed abundance", y = "Observed Reappearance Counts")

ggplot(new.data) +
  geom_point(aes(x = median.abundance, y = pred.m1), color="sienna") +
  labs(x = "OTU median clr-transformed abundance", y = "Simulated Reappearance Counts (model 1)") +
  ylim(0, max(mag.reapp.f4$y))

ggplot(new.data) +
  geom_point(aes(x = median.abundance, y = pred.m2), color="gold") +
  labs(x = "OTU median clr-transformed abundance", y = "Simulated Reappearance Counts (model 2)") +
  ylim(0, max(mag.reapp.f4$y))

ggplot(new.data) +
  geom_point(aes(x = median.abundance, y = pred.m3), color="seagreen") +
  labs(x = "OTU median clr-transformed abundance", y = "Simulated Reappearance Counts (model 3)") +
  ylim(0, max(mag.reapp.f4$y))

ggplot(new.data) +
  geom_point(aes(x = median.abundance, y = pred.m4), color="turquoise") +
  labs(x = "OTU median clr-transformed abundance", y = "Simulated Reappearance Counts (model 4)") +
  ylim(0, max(mag.reapp.f4$y))


ggplot(new.data) +
  geom_point(aes(x = median.abundance, y = pred.m5), color="tomato") +
  labs(x = "OTU median clr-transformed abundance", y = "Simulated Reappearance Counts (model 5)") +
  ylim(0, max(mag.reapp.f4$y))


as.data.frame(table(mag.reapp.f4$y))
as.data.frame(table(new.data$pred.m1))
as.data.frame(table(new.data$pred.m2))
as.data.frame(table(new.data$pred.m3))
as.data.frame(table(new.data$pred.m4))
as.data.frame(table(new.data$pred.m5))

```


# Magnitude of Reappearance Counts without high outliers

```{r}
# don't include any observed reappearance counts above 1000
nonoutlier.reapp.counts <- subset(mag.reapp.f4, y < 1000)

head(nonoutlier.reapp.counts)
```

```{r}
new.m1 <- update(m1, data = nonoutlier.reapp.counts)
new.m2 <- update(m2, data = nonoutlier.reapp.counts)
new.m3 <- update(m3, data = nonoutlier.reapp.counts)
new.m4 <- update(m4, data = nonoutlier.reapp.counts)
new.m5 <- update(m5, data = nonoutlier.reapp.counts)

new.data$pred.new.m1 <- predict_reappearance(X_count = list(new.data$median.abundance), 
                                         X_zero = list(new.data$median.abundance), 
                                         negbin_model = new.m1, zero_inflated = TRUE, seed=0)

new.data$pred.new.m2 <- predict_reappearance(X_count = list(new.data$median.abundance), 
                                         negbin_model = new.m2, zero_inflated = FALSE, seed=0)

new.data$pred.new.m3 <- predict_reappearance(X_count = list(new.data$median.abundance, new.data$time.diff), 
                                         negbin_model = new.m3, zero_inflated = TRUE, seed=0)

new.data$pred.new.m4 <- predict_reappearance(X_count = list(new.data$median.abundance, log(new.data$sample.read.depth)),
                                         negbin_model = new.m4, zero_inflated = TRUE, seed=0)

new.data$pred.new.m5 <- predict_reappearance(X_count = list(new.data$median.abundance, log(new.data$sample.read.depth), new.data$time.diff),
                                         negbin_model = new.m5, zero_inflated = FALSE, seed=0)

ggplot(new.data) +
  geom_point(aes(x = median.abundance, y = pred.new.m1), color="sienna") +
  labs(x = "OTU median clr-transformed abundance", y = "Simulated Reappearance Counts (new model 1)") +
  ylim(0, max(mag.reapp.f4$y))

ggplot(new.data) +
  geom_point(aes(x = median.abundance, y = pred.new.m2), color="gold") +
  labs(x = "OTU median clr-transformed abundance", y = "Simulated Reappearance Counts (new model 2)") +
  ylim(0, max(mag.reapp.f4$y))

ggplot(new.data) +
  geom_point(aes(x = median.abundance, y = pred.new.m3), color="seagreen") +
  labs(x = "OTU median clr-transformed abundance", y = "Simulated Reappearance Counts (new model 3)") +
  ylim(0, max(mag.reapp.f4$y))

ggplot(new.data) +
  geom_point(aes(x = median.abundance, y = pred.new.m4), color="turquoise") +
  labs(x = "OTU median clr-transformed abundance", y = "Simulated Reappearance Counts (new model 4)") +
  ylim(0, max(mag.reapp.f4$y))


ggplot(new.data) +
  geom_point(aes(x = median.abundance, y = pred.new.m5), color="tomato") +
  labs(x = "OTU median clr-transformed abundance", y = "Simulated Reappearance Counts (new model 5)") +
  ylim(0, max(mag.reapp.f4$y))

as.data.frame(table(new.data$pred.new.m1))
as.data.frame(table(new.data$pred.new.m2))
as.data.frame(table(new.data$pred.new.m3))
as.data.frame(table(new.data$pred.new.m4))
as.data.frame(table(new.data$pred.new.m5))

```


```{r}
ggplot(new.data) +
  geom_histogram(aes(pred.m5), binwidth = 1)

ggplot(new.data) +
  geom_histogram(aes(pred.new.m5), binwidth = 1)

```


# Predict Reappearance Relative Abundances

```{r}
reapp.relabs <- format_reappearance_data(otutab = otu.relabs.f4, 
                                         t.sample = f4.metadata$days_since_experiment_start, 
                                         otu.median.abs = median.otu.clrabs.f4,
                                         read.depths = coverage.f4,
                                         otu_id = names(median.otu.clrabs.f4))

head(reapp.relabs); tail(reapp.relabs); summary(reapp.relabs); dim(reapp.relabs)

reapp.relabs
```


The large raw counts are coming from samples with large read counts, which is why we're seeing a similar distribution of reappearing relative abundances. I think the best model to use is a zero-inflated negative binomial regression $y \sim x_1 + x_2$ where $x_1$ is the within-subject OTU median abundance and $x_2$ is log(sample read depth). Or a non-zero-inflated negative binomial regression regressing the count on within-subject OTU median abundance, log(sample read depth), and time since last sample. 

```{r}
ggplot(reapp.relabs) +
  geom_point(aes(x = median.abundance, y = y)) +
  labs(y = "Reappeared Relative Abundances")

ggplot(reapp.relabs) +
  geom_point(aes(x = sample.read.depth, y = y)) +
  labs(y = "Reappeared Relative Abundances")

d <- subset(reapp.relabs, y < 0.04)

ggplot(d) +
  geom_point(aes(x = median.abundance, y = y)) +
  labs(y = "Reappeared Relative Abundances")

ggplot(d) +
  geom_point(aes(x = sample.read.depth, y = y)) +
  labs(y = "Reappeared Relative Abundances")

```



2) Ordinal logistic regression

```{r}
positive.reappearances.f4 <- subset(mag.reapp.f4, y > 0)

positive.y <- positive.reappearances.f4$y
# singeton
# doubleton
# 3 or 4
# 5 or greater
positive.reappearances.f4$y.factor <- factor(case_when(positive.y == 1 ~ 1,
                                                       positive.y == 2 ~ 2,
                                                       positive.y == 3 | positive.y == 4 ~ 3,
                                                       positive.y >= 5 & positive.y <= 10 ~ 5,
                                                       positive.y > 10 ~ 11
                                                       ), 
                                             ordered = TRUE,
                                             levels = c(1, 2, 3, 5, 11))
as.data.frame(table(positive.reappearances.f4$y.factor))

```


The magnitude of reappearance looks associated the OTU's median clr-transformed abundance and the sample's log total read count.
```{r}
ggplot(positive.reappearances.f4, aes(x = median.abundance, y = y.factor , fill = y.factor)) +
  geom_boxplot() +
  labs(x = "OTU median clr-transformed abundance", y = "Reappearance Count Class")

ggplot(positive.reappearances.f4, aes(x = log(sample.read.depth), y = y.factor , fill = y.factor)) +
  geom_boxplot() +
  labs(x = "log Sample Read Depth", y = "Reappearance Count Class")

ggplot(positive.reappearances.f4, aes(x = time.diff, y = y.factor , fill = y.factor)) +
  geom_boxplot() + 
  labs(x = "Difference in Time between Reappearance and Previous Sample", y = "Reappearance Count Class")
```



```{r}
m.olr <- polr(y.factor ~ median.abundance + log(sample.read.depth), 
              data = positive.reappearances.f4, Hess = TRUE)

summary(m.olr)
```


$logit(\hat{P}(Y \leq 1)) = 11.9915 - 0.7849*med.abund - 1.2671*log(read.depth)$
$logit(\hat{P}(Y \leq 2)) = 12.9508 - 0.7849*med.abund - 1.2671*log(read.depth)$
$logit(\hat{P}(Y \leq 3)) = 14.0641 - 0.7849*med.abund - 1.2671*log(read.depth)$

Approximate the t-distribution with a standard normal (good for large sample sizes) and calculate p-values for the coefficients. All the coefficients have significant p-values.
```{r}
olr.ctable <- coef(summary(m.olr))

olr.pvals <- pnorm(abs(olr.ctable[ , "t value"]), lower.tail = FALSE) * 2
(olr.ctable <- cbind(olr.ctable, "p value" = olr.pvals))
```


Approximate CIs with profiling the likelihood function and by assuming a normal distribution. Neither method gives a CI for a coefficient that contains 0.
```{r}
(olr.profiled.ci <- confint(m.olr)) # profiled CIs
confint.default(m.olr) # assuming normality
```

By exponentiating the coefficients, we obtain odds ratios.

For every one unit increase in median clr-transformed abundance, the odds of the reappearance count being greater is multiplied by 2.192, holding the log sample read depth constant.

```{r}
exp(cbind(oddsRatio = coef(m.olr), olr.profiled.ci))
```

Yeah, not a great prediction. Very 1-heavy.
```{r}
summary(predict(m.olr, positive.reappearances.f4))
```

What about an interaction betwee median abundance and log sample read depth? 
median abundance is not so significant a predictor.
Again, the predictions are not great.
```{r}
olr.interact <- polr(y.factor ~ median.abundance*log(sample.read.depth), 
                     data = positive.reappearances.f4, Hess = TRUE)

summary(olr.interact)

olr.interact.ctable <- coef(summary(olr.interact))

olr.interact.pvals <- pnorm(abs(olr.interact.ctable[ , "t value"]), lower.tail = FALSE) * 2
(olr.interact.ctable <- cbind(olr.interact.ctable, "p value" = olr.interact.pvals))
(olr.interact.profiled.ci <- confint(olr.interact)) # profiled CIs
confint.default(olr.interact) # assuming normality

summary(predict(olr.interact, positive.reappearances.f4))
```





