---
title: "Moving Pictures Probability of Reappearance"
author: "Daniel J. Park"
date: "7/17/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(zoib)
library(ZIBR)
library(boot)
require(foreign)
require(MASS)
require(Hmisc)
source("../mp_analysis_functions.R")
```

```{r}
# f4.metadata : metadata sorted by time
# otu.counts.f4 : OTU counts
# otu.relabs.f4 : OTU relative abundances
# otu.clrabs.f4 : OTU clr-transformed abundances
# median.otu.relabs.f4 : OTU's median relative abundance across all samples
# median.otu.clrabs.f4 : OTU's median clr-transformed abundance across all samples
# coverage.f4 : the total number of reads by sample
load("../mp_F4_data.Rdata")
load("../mp_M3_data.Rdata")
load("../dethlefsen_relman.Rdata")
```

# Probability of reappearance

```{r}
big.reapp.tab <- rbind(asin.reapp.tab(dr.D.otutab, 'D'),
                       asin.reapp.tab(dr.E.otutab, 'E'),
                       asin.reapp.tab(dr.F.otutab, 'F'),
                       asin.reapp.tab(otu.counts.f4, 'F4'),
                       asin.reapp.tab(otu.counts.m3, 'M3'))

reapp.plot <- ggplot(big.reapp.tab, aes(x = avg.asin, y = prob.reappear)) +
  geom_point(aes(color = subj.ID),
             size=.8) +
  labs(y = "Probability of colonization",
       x = "OTU mean arcsin Hellinger abundance",
       title = "Probability of colonization by OTU",
       subtitle = "Caporaso et al. / Dethlefsen and Relman",
       color = "Subject ID")

reapp.plot
```

```{r}
big.reapp.tab <- big.reapp.tab %>% 
  mutate(subj.ID = as.factor(subj.ID)) %>%
  drop_na()

ggplot(big.reapp.tab, aes(x = prob.reappear, color = subj.ID)) +
  geom_density(aes(fill = subj.ID),
               alpha = 0.1) + 
  labs(title = "Distribution of conditional reappearance probabilities",
       x = "conditional reappearance probability") +
  scale_fill_discrete()
```

# Predict reappearance with logistic regression GLMM

```{r}
dr.D.avg.asin <- otu.avg.asin(dr.D.otutab)
dr.E.avg.asin <- otu.avg.asin(dr.E.otutab)
dr.F.avg.asin <- otu.avg.asin(dr.F.otutab)
mp.F4.avg.asin <- otu.avg.asin(otu.counts.f4)
mp.M3.avg.asin <- otu.avg.asin(otu.counts.m3)

dr.D.reapp.tab <- format_reappearance_data(dr.D.rel.otutab,
                                           dr.D.avg.asin,
                                           subjID = "D")
dr.E.reapp.tab <- format_reappearance_data(dr.E.rel.otutab,
                                           dr.E.avg.asin,
                                           subjID = "E")
dr.F.reapp.tab <- format_reappearance_data(dr.F.rel.otutab,
                                           dr.F.avg.asin, 
                                           subjID = "F")
mp.F4.reapp.tab <- format_reappearance_data(otu.relabs.f4,
                                            mp.F4.avg.asin,
                                            subjID = "F4")
mp.M3.reapp.tab <- format_reappearance_data(otu.relabs.m3,
                                            mp.M3.avg.asin,
                                            subjID = "M3")

dr.mp.reapp.tab <- rbind(dr.D.reapp.tab,
                         dr.E.reapp.tab,
                         dr.F.reapp.tab,
                         mp.F4.reapp.tab,
                         mp.M3.reapp.tab)

dr.mp.reapp.tab <- dr.mp.reapp.tab %>%
  mutate(reappeared = if_else(y > 0, 1, 0))

head(dr.mp.reapp.tab)
tail(dr.mp.reapp.tab)
dim(dr.mp.reapp.tab)
# save(dr.mp.reapp.tab, file = "../colonization_analysis/bigReappTab.Rdata")
```

```{r}
library(lme4)

reapp.glmm <- glmer(reappeared ~ 1 + (1 | subjID / otu.id), 
                    data = dr.mp.reapp.tab,
                    family = binomial)

summary(reapp.glmm)
plot(reapp.glmm)

reapp.predicted <- predict(reapp.glmm, 
                           re.form = ~ (1 | subjID), 
                           type = "response")
unique(reapp.predicted)

```

```{r}
full.reapp.pred <- predict(reapp.glmm, 
                           re.form = ~ (1 | subjID / otu.id), 
                           type = "response")



full.reapp.pred.df <- data.frame(pred.reapp.prob = full.reapp.pred,
                                 subjID = dr.mp.reapp.tab$subjID)

ggplot(full.reapp.pred.df, aes(x = pred.reapp.prob, color = subjID)) +
  geom_density(aes(fill = subjID),
               alpha = 0.1) + 
  labs(title = "Distribution of predicted reappearance probabilities",
       x = "predicted reappearance probability") +
  xlim(0,1) +
  scale_fill_discrete()

```


# Predicting probability of reappearance with Beta regression 

```{r}
# y ~ beta mean | beta dispersion | 
dat1 <- big.reapp.tab %>%
  filter(subj.ID %in% c("D", "E", "F"))

dat2 <- big.reapp.tab %>%
  filter(prob.reappear == 0 | prob.reappear == 1)
head(dat2)

# beta.glmm1 <- zoib(prob.reappear ~ 1 | 1 | subj.ID,
#               data = dat1,
#               n.response = 1,
#               joint = FALSE,
#               zero.inflation = FALSE,
#               one.inflation = FALSE,
#               EUID = 1:nrow(dat1),
#               random = 1
#               )

library(nlme)
library(lme4)

logit.prob.reappear <- logit(big.reapp.tab$prob.reappear)
summary(logit.prob.reappear)


# reapp.lmm1 <- lmer(formula = logit(prob.reappear) ~ 1 + (1 | subj.ID),
#      data = big.reapp.tab, na.action = "na.omit")
```



```{r}
# a vector of probabilities of reappearing from absence by OTU
prob.reappear.f4 <- reappearance_probabilities(otu.relabs.f4)
prob.reappear.m3 <- reappearance_probabilities(otu.relabs.m3)

# bind in a dataframe with OTUs' median clr-transformed abundance across all samples
reappear.df.f4 <- data.frame(prob.reappear = prob.reappear.f4, 
                             median.clr = median.otu.clrabs.f4)
reappear.df.m3 <- data.frame(prob.reappear = prob.reappear.m3,
                             median.clr = median.otu.clrabs.m3)

reappear.df.f4 <- subset(reappear.df.f4, is.finite(prob.reappear))
reappear.df.m3 <- subset(reappear.df.m3, is.finite(prob.reappear))

reapp.plot.f4 <- ggplot(reappear.df.f4, aes(x = median.clr, y = prob.reappear)) +
  geom_point(color="seagreen") + 
  labs(y = "probability of reappearance", x = "median clr-transformed abundance", title = "OTUs for subject F4")
reapp.plot.m3 <- ggplot(reappear.df.m3, aes(x = median.clr, y = prob.reappear)) +
  geom_point(color="purple") + 
  labs(y = "probability of reappearance", x = "median clr-transformed abundance", title = "OTUs for subject M3")

reapp.plot.f4
reapp.plot.m3
```

# One-inflated Beta Regression
Model the probability of reappearance as a one-inflated beta regression. The mean and dispersion parameters of the beta distribution and  the probability of one-inflation depend on the median clr-transformed abundance.

```{r}
oibr.reapp.f4 <- zoib(prob.reappear ~ median.clr | median.clr | median.clr, data = reappear.df.f4, 
                      joint = FALSE, zero.inflation = FALSE, one.inflation = TRUE, 
                      EUID = rownames(reappear.df.f4), n.iter = 2000)

oibr.reapp.m3 <- zoib(prob.reappear ~ median.clr | median.clr | median.clr, data = reappear.df.m3, 
                      joint = FALSE, zero.inflation = FALSE, one.inflation = TRUE, 
                      EUID = rownames(reappear.df.f4), n.iter = 2000)

save(oibr.reapp.f4, file = "oibr_reapp_f4.Rdata")
save(oibr.reapp.m3, file = "oibr_reapp_m3.Rdata")

```

The potential scale reduction factors from the Gelman plots for both subjects suggest good MCMC convergence. 

```{r}
#load("oibr_reapp_f4.Rdata")
#load("oibr_reapp_m3.Rdata")

traceplot(oibr.reapp.f4$coeff) ; autocorr.plot(oibr.reapp.f4$coeff) ; gelman.diag(oibr.reapp.f4$coeff)
traceplot(oibr.reapp.m3$coeff) ; autocorr.plot(oibr.reapp.m3$coeff) ; gelman.diag(oibr.reapp.m3$coeff)
```

## Simulating probability of colonization by OTU
Coefficients are presented in the order of b, b0, b1, d, where b is for the beta mean, b0 and b1 are for the 0 and 1 inflation, and d is for the beta dispersion.

```{r}
oibr.reapp.coefs.f4 <- summary(oibr.reapp.f4$coeff)
oibr.reapp.coefs.m3 <- summary(oibr.reapp.m3$coeff)

reappear.df.f4$oibr.pred <- sim.zoib(oibr.reapp.f4, 
                                     beta.mean.coefs = oibr.reapp.coefs.f4$statistics[1:2, 1],
                                     infl.coefs = oibr.reapp.coefs.f4$statistics[3:4, 1],
                                     beta.disp.coefs = oibr.reapp.coefs.f4$statistics[5:6, 1],
                                     infl = 1)
reappear.df.m3$oibr.pred <- sim.zoib(oibr.reapp.m3,
                                     beta.mean.coefs = oibr.reapp.coefs.m3$statistics[1:2, 1],
                                     infl.coefs = oibr.reapp.coefs.m3$statistics[3:4, 1],
                                     beta.disp.coefs = oibr.reapp.coefs.m3$statistics[5:6, 1],
                                     infl = 1)
```

## Plot of simulated probability of colonization by OTU

```{r}
require(gridExtra)

sim.reapp.f4 <- ggplot(reappear.df.f4, aes(x = median.clr, y = oibr.pred)) +
  geom_point(color="sienna") + 
  labs(y = "OIBR simulated probability of reappearance", x = "median clr-transformed abundance", title = "OTUs for subject F4")
sim.reapp.m3 <- ggplot(reappear.df.m3, aes(x = median.clr, y = oibr.pred)) +
  geom_point(color="turquoise") + 
  labs(y = "OIBR simulated probability of reappearance", x = "median clr-transformed abundance", title = "OTUs for subject M3")

grid.arrange(reapp.plot.f4, sim.reapp.f4, nrow = 1)
grid.arrange(reapp.plot.m3, sim.reapp.m3, nrow = 1)

```

